Live Functional Programming with Typed Holes

### 2018-03-28 calculus.tex
remove the H- prefix and a few minor wording tweaks to 3.1

The syntax of the core calculus, specified in Fig.~\ref{fig:hazelnut-live-syntax}, consists of types and expressions with holes, which we refer to as H-types and H-expressions, respectively. We distinguish between \emph{external} H-expressions, $e$, and \emph{internal} H-expressions, $d$. External H-expressions correspond to programs as entered by the programmer (see Sec.~\ref{sec:implementation}\todo{or maybe Sec. 2?}{} for examples of manual, semi-automated and fully automated hole entry methods). Each well-typed external H-expression (as specified in Sec.~\ref{sec:external-statics}) expands to a well-typed internal H-expression (Sec.~\ref{sec:expansion}) before it is evaluated (Sec.~\ref{sec:evaluation}). We distinguish the external and internal languages because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is  formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics\todo{cite frank's notes}; and (2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts. This machinery is initialized by the expansion step, rather than entered explicitly by the programmer. In this regard, the internal language is analagous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects.

@@

Let us start with a brief overview of the static semantics of the \HazelnutLive external language. H-types, $\htau$, classify external H-expressions, $e$, according to the type system in Fig. \ref{fig:bidirectional-typing}, which closely follows the \Hazelnut type system \cite{popl-paper} (the minor differences are discussed as they come up). The type system is specified in the \emph{bidirectional} style \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da} around two mutually defined judgements. The type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, synthesizes a type $\htau$ for $\hexp$ assuming $\hGamma$, which tracks typing assumptions of the form $x : \htau$ in the usual manner. The type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, instead checks $\hexp$ against a given $\htau$. Algorithmically, the type is an input of type analysis.% Algorithmically, the type is an output of type synthesis but an input of type analysis.

@@

The matched arrow type judgement also appears in the rule governing function application, Rule {SAp}, which treats expressions in function position of hole type as if they were of type $\tarr{\tehole}{\tehole}$.

@@

\autoref{fig:complete} defines \emph{complete} H-types, H-expressions, and
dynamic H-expressions to be those terms without type or expression holes.

>>>>

The syntax of the core calculus, specified in Fig.~\ref{fig:hazelnut-live-syntax}, consists of types and expressions with holes. We distinguish between \emph{external} expressions, $e$, and \emph{internal} expressions, $d$. External expressions correspond to programs as entered by the programmer (see Sec.~\ref{sec:implementation}\todo{or maybe Sec. 2?}{} for examples of manual, semi-automated and fully automated hole entry methods). Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics}) expands to a well-typed internal expression (Sec.~\ref{sec:expansion}) before it is evaluated (Sec.~\ref{sec:evaluation}). We distinguish the external and internal languages because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is  formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics\todo{cite frank's notes}; and (2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts. This machinery is inserted by the expansion step, rather than entered explicitly by the programmer. In this regard, the internal language is analagous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects.

@@

Let us start with a brief overview of the static semantics of the \HazelnutLive external language. Types, $\htau$, classify external expressions, $e$, according to the type system in Fig. \ref{fig:bidirectional-typing}, which closely follows the \Hazelnut type system \cite{popl-paper} (the minor differences are discussed as they come up). It is specified in the \emph{bidirectional} style \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da} around two mutually defined judgements. The type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, synthesizes a type $\htau$ for $\hexp$ assuming $\hGamma$, which tracks typing assumptions of the form $x : \htau$ in the usual manner. The type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, instead checks $\hexp$ against a given $\htau$. Algorithmically, the type is an input of type analysis.% Algorithmically, the type is an output of type synthesis but an input of type analysis.

@@

The rule governing function application, Rule {SAp}, similarly treats expressions in function position of hole type as if they were of type $\tarr{\tehole}{\tehole}$ using the matched arrow type judgement.

@@

\autoref{fig:complete} defines \emph{complete} types, expressions, and
dynamic expressions to be those terms without type or expression holes.
